\pkg{sparseHessianFD}: An \proglang{R} Package for Estimating Sparse Hessian Matrices

### 2018-03-26 nobuild/jss/2717/paper.tex

 \documentclass[jss]{jss}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\usepackage{Sweave}

%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{sparseHessianFD}
%\VignettePackage{sparseHessianFD}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{tabularx}
\usepackage{dcolumn}
\usepackage{booktabs}
\usepackage{etoolbox}
\usepackage{placeins}
\newtoggle{tikz}
\toggletrue{tikz}

\iftoggle{tikz}{
\usepackage{tikz}
}{}

 \newcolumntype{d}[1]{D{.}{.}{#1}}
\newcommand{\func}[1]{\code{#1}}
\newcommand{\class}[1]{\textsl{#1}}
\newcommand{\funcarg}[1]{\code{#1}}
\newcommand{\variable}[1]{\code{#1}}
\newcommand{\method}[1]{\func{#1}}

 \newcommand{\df}[3]{\mathsf{d}^{#1}f(#2;#3)}
 \newcommand{\parD}[3]{\mathsf{D}^{#1}_{#2}#3}
\newcommand{\hess}[2]{\mathsf{H}_{#1}#2}
\newcommand{\hessLT}[2]{\mathsf{L}_{#1}#2}
\newcommand{\Mat}[1]{\mathbf{#1}}
\newcommand{\Real}[1]{\mathbb{R}^{#1}}
\newcommand{\Complex}[1]{\mathbb{Z}^{#1}}
\DeclareMathOperator{\Imag}{Im}

\author{Michael Braun\\Edwin L. Cox School of Business\\Southern Methodist University}
\Plainauthor{Michael Braun}
\title{\pkg{sparseHessianFD}: An R Package for Estimating Sparse Hessian Matrices}
\Plaintitle{sparseHessianFD: An R Package for Estimating Sparse Hessian Matrices}
\Shorttitle{sparseHessianFD:  Sparse Hessians}
\Keywords{sparse Hessians, sparsity, computation of Hessians,
  graph coloring, finite differences, differentiation, complex step}
\Address{
  Michael Braun\\
  Edwin L. Cox School of Business\\
  Southern Methodist University\\
  6212 Bishop Blvd.\\
  Dallas, TX 75275\\
  E-mail:  \email{braunm@smu.edu}\\
  URL:  \url{http://www.smu.edu/Cox/Departments/FacultyDirectory/BraunMichael}
}

\Abstract{
  Sparse Hessian matrices occur often
  in statistics, and their fast and accurate estimation can
  improve efficiency of numerical optimization and sampling
  algorithms. By exploiting the known sparsity pattern of a Hessian, methods in the
  sparseHessianFD package require many fewer function or gradient
  evaluations than would be required if the Hessian were treated as
  dense. The package implements established graph coloring and
  linear substitution algorithms that were previously unavailable to
  R users, and is most useful when other numerical,
  symbolic or algorithmic methods are impractical, inefficient or
  unavailable.}

\date{January 25, 2017}

%% need no \usepackage{Sweave.sty}

%%\newenvironment{knitrout}{\begin{figure}}{\end{figure}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

The Hessian of an objective function with $M$ variables has $M^2$
elements, of which $M(M+1)/2$ are unique.  Thus, the storage
requirements of the Hessian, and computational cost of many linear
algebra operations on it, grow quadratically with the number of
decision variables.   For
applications with hundreds of thousands of variables, computing the
Hessian even once might not be practical under time, storage or
processor constraints.  Hierarchical models, in
which each additional heterogeneous unit is associated with its own subset of
variables, are particularly vulnerable to this curse of dimensionality

However, for many problems, the Hessian is
\emph{sparse}, meaning that the proportion of  ``structural''  zeros
(matrix elements that are always zero, regardless of the value at which
function is estimated) is high.  Consider a log
posterior density in a Bayesian hierarchical
model.  If the outcomes across units are conditionally
independent, the cross-partial derivatives of heterogeneous variables
across units are zero.  As the number of units
increases, the size of the Hessian still grows quadratically, but the number
of \emph{non-zero} elements grows only linearly; the Hessian
becomes increasingly sparse.  The row and column indices of the
non-zero elements comprise the \emph{sparsity pattern} of the Hessian,
and are typically known in advance, before computing the values of
those elements.  \proglang{R} packages such as \pkg{trustOptim}
\citep{R_trustOptim}, \pkg{sparseMVN} \citep{R_sparseMVN} and
\pkg{ipoptr} \citep{R_ipoptr} have the capability to accept Hessians
in a compressed sparse format.

The \pkg{sparseHessianFD} package is a tool for estimating
sparse Hessians numerically, using either finite differences or
complex perturbations of
gradients.  Section~\ref{sec:numdiff} will
cover the specifics, but the basic idea is as follows.  Consider a real-valued
function $f(x)$, its gradient $\nabla{f(x)}$, and its Hessian
$\hess{}{f(x)}$, for $x\in\mathbb{R}^M$.  Define the derivative vector
as the transpose of the gradient, and a vector of partial derivatives, so
${\parD{}{}{f(x)}=\nabla f(x)^\top=\left(\parD{}{1}{},\mathellipsis,\parD{}{M}{}\right)}$.   (Throughout the paper, we will try
to reduce notational clutter by referring to the derivative and
Hessian as $\parD{}{}{}$ and $\hess{}{}$, respectively, without the
$f(x)$ symbol).  Let $e_m$ be a vector of zeros, except with a 1 in the $m$th
element, and let $\delta$ be a sufficiently small scalar
constant. A ``finite difference'' linear approximation to the $m$th column of the Hessian is ${\hess{m}{}\approx\left(\nabla f(x+\delta
  e_m)- \nabla f(x)\right)/\delta}$.  Estimating a dense Hessian in this way involves at least $M+1$ calculations of the
gradient: one for the gradient at $x$, and one after perturbing each
of the $M$ elements of $x$, one at a time. Under certain conditions, a
more accurate approximation is the ``complex step'' method:  $\hess{m}{}\approx\Imag\left(\nabla f(x+i\delta
  e_m)\right)/\delta$, where $i=\sqrt{-1}$ and $\Imag$ returns the
imaginary part of a complex
number \citep{SquireTrapp1998}.  Regardless of the approximation method used, if the Hessian is sparse, most of the
elements are constrained to zero.  Depending
on the sparsity pattern of the Hessian, those constraints may let us recover
the Hessian with fewer gradient evaluations by perturbing multiple
elements of $x$ together.  For some sparsity patterns, estimating a Hessian in
this way can be profoundly efficient.  In fact, for the hierarchical models that we consider in
this paper, the number of gradient evaluations does not increase with
the number of additional heterogeneous units.

@@

As with any computing method or algorithm, there are boundaries around
the space of applications for which \pkg{sparseHessianFD} is the right
tool for the job. In general, numerical approximations are not ``first
choice'' methods because the result is not exact, so
\pkg{sparseHessianFD} should not be used when the application cannot
tolerate any error, no matter how small.  Also, we admit that some
users might balk at having to provide an exact gradient, even though
the Hessian will be estimated numerically.\footnote{It is possible
  that a gradient that is approximated using a complex step method
  would be precise enough to be useful. A further investigation is
  beyond the scope of this paper.} However, deriving a vector of first
derivatives, and writing \proglang{R} functions to compute them, is a
lot easier than doing the same for a matrix of second derivatives, and
more accurate than computing second-order approximations from the
objective function. Even when we have derived the Hessian
symbolically, in practice it may still be faster to estimate the Hessian
using \pkg{sparseHessianFD} than coding it directly.  These are the
situations in which \pkg{sparseHessianFD} adds the most value to the
statistician's toolbox.

@@

Before describing how to use the package, we
present two short background notes.  The first note is an informal
mathematical explanation of numerical estimation of the Hessian
matrix, with an illustration of how the number of gradient estimates
can be reduced by exploiting the sparsity pattern and symmetric
structure.  This note borrows heavily from, and uses the notation in,
\citet[Chapter 6]{MagnusNeudecker2007}. The
second note is a summary of some of the sparse matrix classes that are
defined in the \pkg{Matrix} package \citep{R_Matrix}, which are used
extensively in \pkg{sparseHessianFD}.

@@

The partial derivative of a real scalar-valued function $f(x)$ with
respect to $x_j$ (the $j$th
component of $x\in\Real{M}$) is defined as
\begin{align}
  \label{eq:defParD}
\parD{}{j}{f(x)}=\lim\limits_{\delta\to 0}\frac{f(x+\delta e_j)-f(x)}{\delta}
\end{align}
For a sufficiently small $\delta$, this definition allows for a
linear approximation to $\parD{}{j}{f(x)}$.  The derivative of $f(x)$
is the vector of all $M$ partial derivatives.
\begin{align}
  \parD{}{}{f(x)}=\left(\parD{}{1}{f(x)},\mathellipsis,\parD{}{M}{f(x)}\right)
  \end{align}
 The \emph{gradient} is defined as $\nabla f(x)=\parD{}{}{f(x)}^\top$.

We define the second-order partial derivative as
\begin{align}
  \label{eq:14}
  \parD{2}{jk}{}=\lim\limits_{\delta\to 0}\frac{\parD{}{j}{f(x+\delta e_k)}-\parD{}{j}{f(x)}}{\delta}
\end{align}
and the Hessian as
\begin{align}
  \label{eq:15}
  \hess{}{f(x)}=
  \begin{pmatrix}
    \parD{2}{11}{}&  \parD{2}{12}{}&  \mathellipsis &  \parD{2}{1M}{}\\
    \parD{2}{21}{}&  \parD{2}{22}{}&  \mathellipsis &  \parD{2}{2M}{}\\
    \vdots&\vdots&&\vdots\\
    \parD{2}{M1}{}&  \parD{2}{M2}{}&  \mathellipsis &  \parD{2}{MM}{}
    \end{pmatrix}
\end{align}

The Hessian is symmetric, so $\parD{2}{ij}{}=\parD{2}{ji}{}$.

@@

To estimate the $m$th column of $\hess{}{}$ using finite differences, we choose a
sufficiently small $\delta$, and compute
\begin{align}
  \label{eq:1}
  \hess{m}{f(x)}&\approx\frac{\parD{}{}{f(x+\delta e_m)}-\parD{}{}{f(x)}}{\delta}
\end{align}

For $M=2$, our estimate of a general $\hess{}{f(x)}$ would be
\begin{align}
  \label{eq:FDhess2}
  \hess{}{f(x)}&\approx
  \begin{pmatrix}
    \parD{}{1}{f(x_1+\delta, x_2)}-\parD{}{1}{f(x_1,x_2)}& \parD{}{1}{f(x_1,x_2+\delta)}-\parD{}{1}{f(x_1,x_2)}\\
        \parD{}{2}{f(x_1+\delta, x_2)}-\parD{}{2}{f(x_1,x_2)}&  \parD{}{2}{f(x_1,x_2+\delta)}-\parD{}{2}{f(x_1,x_2)}
    \end{pmatrix}/\delta
\end{align}

This estimate requires three evaluations of the gradient to get
${\parD{}{}{f(x_1,x_2)}}$, ${\parD{}{}{f(x_1+\delta,x_2)}}$, and
${\parD{}{}{f(x_1,x_2+\delta)}}$.

Now suppose that the Hessian is sparse, and that the off-diagonal
elements are zero.  That means that
\begin{align}
  \parD{}{1}{f(x_1,x_2+\delta)}-\parD{}{1}{f(x_1,x_2)}&=0\label{eq:I1}\\
    \parD{}{2}{f(x_1+\delta,x_2)}-\parD{}{2}{f(x_1,x_2)}&=0\label{eq:I2}
\end{align}
If the identity in Equation~\ref{eq:I1} holds for $x_1$, it must also
hold for $x_1+\delta$, and if Equation~\ref{eq:I2} holds for $x_2$, it
must also hold for $x_2+\delta$.  Therefore,
\begin{align}
  \label{eq:FDhess2sp}
  \hess{}{f(x)}&\approx
  \begin{pmatrix}
    \parD{}{1}{f(x_1+\delta, x_2+\delta)}-\parD{}{1}{f(x_1,x_2)}&0\\
        0&  \parD{}{2}{f(x_1+\delta,x_2+\delta)}-\parD{}{2}{f(x_1,x_2)}
    \end{pmatrix}/\delta
\end{align}

Only two gradients,
${\parD{}{}{f(x_1,x_2)}}$ and ${\parD{}{}{f(x_1+\delta,x_2+\delta)}}$,
are needed.
Being able to reduce the number of gradient evaluations from 3 to 2 depends on knowing that
the cross-partial derivatives are zero.

@@

If $f(x)$ is defined over a complex domain and is holomorphic, then we
can we can approximate $\parD{}{}{f(x)}$
and $\hess{}{f(x)}$ at real values of $x$ using the complex step
method. This method comes from a Taylor
series expansion of $f(x)$ in the imaginary direction of the complex
plane \citep{SquireTrapp1998}.  After rearranging terms, and taking the imaginary
parts of both sides,
\begin{align}
  \label{eq:complexTaylor}
  f(x+i\delta)&=f(x) + i\delta\parD{}{}{f(x)}+\mathcal{O}(\delta^2)\\
  \parD{}{}{f(x)}&\approx\frac{\Imag(f(x+i\delta))}{\delta}
\end{align}

Estimating a first derivative using the complex step method does not
require a differencing operation, so there is no subtraction operation
that might generate roundoff errors.  Thus, the approximation can be
made arbitrarily precise as $\delta\rightarrow 0$ \citep{LaiCrassidis2008}.  This is not the
case for second-order approximations of
the Hessian \citep{AbreuStich2013}.  However, when the gradient can be
computed exactly, we can compute a first-order approximation to
Hessian by treating it as the Jacobian of
a vector-valued function \citep{LaiCrassidis2008}.

\begin{align}
  \label{eq:CShess2}
  \hess{}{f(x)}&\approx
  \begin{pmatrix}
    \Imag(\parD{}{1}{f(x_1+i\delta, x_2)})& \Imag(\parD{}{1}{f(x_1,x_2+i\delta)})\\
        \Imag(\parD{}{2}{f(x_1+i\delta, x_2)})&\Imag(\parD{}{2}{f(x_1,x_2+i\delta)})
    \end{pmatrix}/\delta
\end{align}

If this matrix were dense, we would need two evaluations of the
$\parD{}{}f(x)$ to estimate it.  If the matrix were sparse, with the same sparsity
pattern as the Hessian in Equation~\ref{eq:FDhess2sp}, and we
assume that structural zeros remain zero for all
complex $x\in\Complex{M}$, then we need only one evaluation.  Suppose
we were to subtract $\Imag(\parD{}{}{f(x_1,x_2)})$ from each column of
$\hess{}{f(x)}$. When $x$ is real, the imaginary part of the gradient
is zero, so this operation has no effect on the value of the
Hessian. But the sparsity constraints ensure that
the following identities hold all complex $x$.

\begin{align}
  \Imag(\parD{}{1}{f(x_1,x_2+i\delta)}) -\Imag(\parD{}{1}{f(x_1,x_2)})&=0\label{eq:I1c}\\
    \Imag(\parD{}{2}{f(x_1+i\delta,x_2)})-\Imag(\parD{}{2}{f(x_1,x_2)})&=0\label{eq:I2c}
\end{align}

As with the finite difference method, because Equation~\ref{eq:I1c} holds
for $x_1$, it must also hold for $x_1+i\delta$, and because Equation~\ref{eq:I2c} holds
for $x_2$, it must also hold for $x_2+i\delta$. Thus,

\begin{align}
  \label{eq:CShess2sp}
  \hess{}{f(x)}&\approx
  \begin{pmatrix}
    \Imag(\parD{}{1}{f(x_1+i\delta, x_2+i\delta)})&0\\
        0&\Imag(\parD{}{2}{f(x_1+i\delta,x_2+i\delta)})
    \end{pmatrix}/\delta
\end{align}

for real $x$.  Only one evaluation of the gradient is required.

@@

Next, define $\Mat{Y}\in\Real{M\times C}$ such that each column is
either a difference in gradients, or the imaginary part of a
complex-valued gradient, depending on the chosen method.
\begin{align}
  \label{eq:Yg}
  \Mat{Y}_c&=
           \begin{cases}
             \nabla f(x+G_c)-\nabla f(x)&\text{finite difference method}\\
             \Imag(\nabla f(x+iG_c))&\text{complex step method}
             \end{cases}
\end{align}

If $C=M$, then $\Mat{G}$ is a diagonal matrix with $\delta$ in each
diagonal element.  The matrix equation $\hess{}{}\Mat{G}=\Mat{Y}$ represents the linear approximation
$\hess{im}{}\delta\approx y_{im}$, and we can solve for all elements of $\hess{}{}$
just by computing $\Mat{Y}$. But if $C<M$,
there must be at least one $\Mat{G}_c$ with $\delta$ in at least two
rows. The corresponding column $\Mat{Y}_c$ is computed by perturbing multiple variables at
once, so we cannot solve for any $\hess{im}{}$
without further constraints.

These constraints come from the sparsity pattern and symmetry
of the Hessian. Consider an example with the following values and
sparsity pattern.
\begin{align}
  \label{eq:7}
 \hess{}{f(x)}= \begin{pmatrix}
    h_{11}&0&h_{31}&0&0\\
    0&h_{22}&0&h_{42}&0\\
    h_{31}&0&h_{33}&0&h_{53}\\
    0&h_{42}&0&h_{44}&0\\
    0&0&h_{53}&0&h_{55}
  \end{pmatrix}
\end{align}

Suppose $C=2$, and define group membership of the five variables through the following $\Mat{G}$ matrix.
\begin{align}
  \label{eq:7}
 \Mat{G}^\top= \begin{pmatrix}
 \delta&\delta&0&0&\delta\\
 0&0&\delta&\delta&0
 \end{pmatrix}
\end{align}
Variables 1, 2 and 5 are in group 1, while variables 3 and 4 are in
group 2.

>>>>

\documentclass[article]{jss}
\usepackage{thumbpdf} 
\graphicspath{{Figures/}}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{tabularx}
\usepackage{tikz}

\newcommand{\func}[1]{\code{#1()}}
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\funcarg}[1]{\code{#1}}
\newcommand{\variable}[1]{\code{#1}}
\newcommand{\method}[1]{\func{#1}}

\newcommand{\parD}[3]{\mathsf{D}^{#1}_{#2}#3}
\newcommand{\hess}[2]{\mathsf{H}_{#1}#2}
\newcommand{\hessLT}[2]{\mathsf{L}_{#1}#2}
\newcommand{\Mat}[1]{\mathbf{#1}}
\newcommand{\Real}[1]{\mathbb{R}^{#1}}
\newcommand{\Complex}[1]{\mathbb{Z}^{#1}}
\DeclareMathOperator{\Imag}{Im}

\author{Michael Braun\\Southern Methodist University}
\Plainauthor{Michael Braun}

\title{\pkg{sparseHessianFD}: An \proglang{R} Package for Estimating Sparse Hessian Matrices}

\Plaintitle{sparseHessianFD: An R Package for Estimating Sparse Hessian Matrices}

\Keywords{sparse Hessians, sparsity, computation of Hessians,
  graph coloring, finite differences, differentiation, complex step}

\Address{
  Michael Braun\\
  Edwin L. Cox School of Business\\
  Southern Methodist University\\
  6212 Bishop Blvd.\\
  Dallas, TX 75275\\
  E-mail:  \email{braunm@smu.edu}\\
  URL:  \url{http://www.smu.edu/Cox/Departments/FacultyDirectory/BraunMichael}
}

\Abstract{
  Sparse Hessian matrices occur often
  in statistics, and their fast and accurate estimation can
  improve efficiency of numerical optimization and sampling
  algorithms. By exploiting the known sparsity pattern of a Hessian, methods in the
  \pkg{sparseHessianFD} package require many fewer function or gradient
  evaluations than would be required if the Hessian were treated as
  dense. The package implements established graph coloring and
  linear substitution algorithms that were previously unavailable to
  \proglang{R} users, and is most useful when other numerical,
  symbolic or algorithmic methods are impractical, inefficient or
  unavailable.}

\begin{document}

@@

The Hessian of an objective function with $M$ variables has $M^2$
elements, of which $M(M+1)/2$ are unique.  Thus, the storage
requirements of the Hessian, and computational cost of many linear
algebra operations on it, grow quadratically with the number of
decision variables.   For
applications with hundreds of thousands of variables, computing the
Hessian even once might not be practical under the given time, storage or
processor constraints.  Hierarchical models, in
which each additional heterogeneous unit is associated with its own subset of
variables, are particularly vulnerable to this curse of dimensionality

However, for many problems, the Hessian is \emph{sparse}, meaning that
the proportion of ``structural'' zeros (matrix elements that are
always zero, regardless of the value at which the function is
estimated) is high.  Consider a log posterior density in a Bayesian
hierarchical model.  If the outcomes across units are conditionally
independent, the cross-partial derivatives of heterogeneous variables
across units are zero.  As the number of units increases, the size of
the Hessian still grows quadratically, but the number of
\emph{non-zero} elements grows only linearly; the Hessian becomes
increasingly sparse.  The row and column indices of the non-zero
elements comprise the \emph{sparsity pattern} of the Hessian, and are
typically known in advance, before computing the values of those
elements.  \proglang{R} \citep{R} packages such as \pkg{trustOptim}
\citep{R_trustOptim}, \pkg{sparseMVN} \citep{R_sparseMVN} and
\pkg{ipoptr} \citep{R_ipoptr} have the capability to accept Hessians
in a compressed sparse format.

The \pkg{sparseHessianFD} package \citep{sparseHessianFD} is a tool
for estimating sparse Hessians numerically, using either finite
differences or complex perturbations of gradients. The package is
available from the Comprehensive \proglang{R} Archive Network (CRAN)
at \url{https://CRAN.R-project.org/package=sparseHessianFD}.
Section~\ref{sec:numdiff} will cover the specifics, but the basic idea
is as follows.  Consider a real-valued function $f(x)$, its gradient
$\nabla{f(x)}$, and its Hessian $\hess{}{f(x)}$, for
$x\in\mathbb{R}^M$.  Define the derivative vector as the transpose of
the gradient, and a vector of partial derivatives, so
${\parD{}{}{f(x)}=\nabla
  f(x)^\top=\left(\parD{}{1}{},\mathellipsis,\parD{}{M}{}\right)}$.
(Throughout the paper, we will try to reduce notational clutter by
referring to the derivative and Hessian as $\parD{}{}{}$ and
$\hess{}{}$, respectively, without the $f(x)$ symbol).  Let $e_m$ be a
vector of zeros, except with a 1 in the $m$th element, and let
$\delta$ be a sufficiently small scalar constant. A ``finite
difference'' linear approximation to the $m$th column of the Hessian
is
${\hess{m}{}\approx\left(\nabla f(x+\delta e_m)- \nabla
    f(x)\right)/\delta}$.  Estimating a dense Hessian in this way
involves at least $M+1$ calculations of the gradient: one for the
gradient at $x$, and one after perturbing each of the $M$ elements of
$x$, one at a time. Under certain conditions, a more accurate
approximation is the ``complex step'' method:
$\hess{m}{}\approx\Imag\left(\nabla f(x+i\delta e_m)\right)/\delta$,
where $i=\sqrt{-1}$ and $\Imag()$ returns the imaginary part of a
complex number \citep{SquireTrapp1998}.  Regardless of the
approximation method used, if the Hessian is sparse, most of the
elements are constrained to zero.  Depending on the sparsity pattern
of the Hessian, those constraints may let us recover the Hessian with
fewer gradient evaluations by perturbing multiple elements of $x$
together.  For some sparsity patterns, estimating a Hessian in this
way can be profoundly efficient.  In fact, for the hierarchical models
that we consider in this paper, the number of gradient evaluations
does not increase with the number of additional heterogeneous units.

@@

As with any computing method or algorithm, there are boundaries around
the space of applications for which package \pkg{sparseHessianFD} is
the right tool for the job. In general, numerical approximations are
not ``first choice'' methods because the result is not exact, so
package \pkg{sparseHessianFD} should not be used when the application
cannot tolerate any error, no matter how small.  Also, we admit that
some users might balk at having to provide an exact gradient, even
though the Hessian will be estimated numerically.\footnote{It is
  possible that a gradient that is approximated using a complex step
  method would be precise enough to be useful. A further investigation
  is beyond the scope of this paper.} However, deriving a vector of
first derivatives, and writing \proglang{R} functions to compute them,
is a lot easier than doing the same for a matrix of second
derivatives, and more accurate than computing second-order
approximations from the objective function. Even when we have derived
the Hessian symbolically, in practice it may still be faster to
estimate the Hessian using package \pkg{sparseHessianFD} than coding
it directly.  These are the situations in which package
\pkg{sparseHessianFD} adds the most value to the statistician's
toolbox.

@@

Before describing how to use the package, we
present two short background notes.  The first note is an informal
mathematical explanation of numerical estimation of the Hessian
matrix, with an illustration of how the number of gradient estimates
can be reduced by exploiting the sparsity pattern and symmetric
structure.  This note borrows heavily from, and uses the notation in,
\citet[Chapter 6]{MagnusNeudecker2007}. The
second note is a summary of some of the sparse matrix classes that are
defined in the \pkg{Matrix} package \citep{R_Matrix}, which are used
extensively in package \pkg{sparseHessianFD}.

@@

The partial derivative of a real scalar-valued function $f(x)$ with
respect to $x_j$ (the $j$th
component of $x\in\Real{M}$) is defined as
\begin{align*}
\parD{}{j}{f(x)}=\lim\limits_{\delta\to 0}\frac{f(x+\delta e_j)-f(x)}{\delta}.
\end{align*}
For a sufficiently small $\delta$, this definition allows for a
linear approximation to $\parD{}{j}{f(x)}$.  The derivative of $f(x)$
is the vector of all $M$ partial derivatives:
\begin{align*}
  \parD{}{}{f(x)}=\left(\parD{}{1}{f(x)},\mathellipsis,\parD{}{M}{f(x)}\right).
\end{align*}
The \emph{gradient} is defined as $\nabla f(x)=\parD{}{}{f(x)}^\top$.

We define the second-order partial derivative as
\begin{align*}
  \parD{2}{jk}{}=\lim\limits_{\delta\to 0}\frac{\parD{}{j}{f(x+\delta e_k)}-\parD{}{j}{f(x)}}{\delta}
\end{align*}
and the Hessian as
\begin{align*}
  \hess{}{f(x)}=
  \begin{pmatrix}
    \parD{2}{11}{}&  \parD{2}{12}{}&  \mathellipsis &  \parD{2}{1M}{}\\
    \parD{2}{21}{}&  \parD{2}{22}{}&  \mathellipsis &  \parD{2}{2M}{}\\
    \vdots&\vdots&&\vdots\\
    \parD{2}{M1}{}&  \parD{2}{M2}{}&  \mathellipsis &  \parD{2}{MM}{}
    \end{pmatrix}.
\end{align*}
The Hessian is symmetric, so $\parD{2}{ij}{}=\parD{2}{ji}{}$.

@@

To estimate the $m$th column of $\hess{}{}$ using finite differences, we choose a
sufficiently small $\delta$, and compute
\begin{align*}
  \hess{m}{f(x)}&\approx\frac{\parD{}{}{f(x+\delta e_m)}-\parD{}{}{f(x)}}{\delta}.
\end{align*}
For $M=2$, our estimate of a general $\hess{}{f(x)}$ would be
\begin{align*}
  \hess{}{f(x)}&\approx
  \begin{pmatrix}
    \parD{}{1}{f(x_1+\delta, x_2)}-\parD{}{1}{f(x_1,x_2)}& \parD{}{1}{f(x_1,x_2+\delta)}-\parD{}{1}{f(x_1,x_2)}\\
        \parD{}{2}{f(x_1+\delta, x_2)}-\parD{}{2}{f(x_1,x_2)}&  \parD{}{2}{f(x_1,x_2+\delta)}-\parD{}{2}{f(x_1,x_2)}
    \end{pmatrix}/\delta.
\end{align*}
This estimate requires three evaluations of the gradient to get
${\parD{}{}{f(x_1,x_2)}}$, ${\parD{}{}{f(x_1+\delta,x_2)}}$, and
${\parD{}{}{f(x_1,x_2+\delta)}}$.

Now suppose that the Hessian is sparse, and that the off-diagonal
elements are zero.  That means that:
\begin{align}
  \parD{}{1}{f(x_1,x_2+\delta)}-\parD{}{1}{f(x_1,x_2)}&=0,\label{eq:I1}\\
    \parD{}{2}{f(x_1+\delta,x_2)}-\parD{}{2}{f(x_1,x_2)}&=0.\label{eq:I2}
\end{align}
If the identity in Equation~\ref{eq:I1} holds for $x_1$, it must also
hold for $x_1+\delta$, and if Equation~\ref{eq:I2} holds for $x_2$, it
must also hold for $x_2+\delta$.  Therefore,
\begin{align}
  \label{eq:FDhess2sp}
  \hess{}{f(x)}&\approx
  \begin{pmatrix}
    \parD{}{1}{f(x_1+\delta, x_2+\delta)}-\parD{}{1}{f(x_1,x_2)}&0\\
        0&  \parD{}{2}{f(x_1+\delta,x_2+\delta)}-\parD{}{2}{f(x_1,x_2)}
    \end{pmatrix}/\delta.
\end{align}
Only two gradients,
${\parD{}{}{f(x_1,x_2)}}$ and ${\parD{}{}{f(x_1+\delta,x_2+\delta)}}$,
are needed.
Being able to reduce the number of gradient evaluations from 3 to 2 depends on knowing that
the cross-partial derivatives are zero.

@@

If $f(x)$ is defined over a complex domain and is holomorphic, then we
can approximate $\parD{}{}{f(x)}$ and $\hess{}{f(x)}$ at real values
of $x$ using the complex step method. This method comes from a Taylor
series expansion of $f(x)$ in the imaginary direction of the complex
plane \citep{SquireTrapp1998}.  After rearranging terms, and taking
the imaginary parts of both sides,
\begin{align*}
  f(x+i\delta)&=f(x) + i\delta\parD{}{}{f(x)}+\mathcal{O}(\delta^2),\\
  \parD{}{}{f(x)}&\approx\frac{\Imag(f(x+i\delta))}{\delta}.
\end{align*}
Estimating a first derivative using the complex step method does not
require a differencing operation, so there is no subtraction operation
that might generate roundoff errors.  Thus, the approximation can be
made arbitrarily precise as $\delta\rightarrow 0$ \citep{LaiCrassidis2008}.  This is not the
case for second-order approximations of
the Hessian \citep{AbreuStich2013}.  However, when the gradient can be
computed exactly, we can compute a first-order approximation to the
Hessian by treating it as the Jacobian of
a vector-valued function \citep{LaiCrassidis2008}:
\begin{align*}
  \hess{}{f(x)}&\approx
  \begin{pmatrix}
    \Imag(\parD{}{1}{f(x_1+i\delta, x_2)})& \Imag(\parD{}{1}{f(x_1,x_2+i\delta)})\\
        \Imag(\parD{}{2}{f(x_1+i\delta, x_2)})&\Imag(\parD{}{2}{f(x_1,x_2+i\delta)})
    \end{pmatrix}/\delta.
\end{align*}
If this matrix were dense, we would need two evaluations of the
$\parD{}{}f(x)$ to estimate it.  If the matrix were sparse, with the same sparsity
pattern as the Hessian in Equation~\ref{eq:FDhess2sp}, and we
assume that structural zeros remain zero for all
complex $x\in\Complex{M}$, then we need only one evaluation.  Suppose
we were to subtract $\Imag(\parD{}{}{f(x_1,x_2)})$ from each column of
$\hess{}{f(x)}$. When $x$ is real, the imaginary part of the gradient
is zero, so this operation has no effect on the value of the
Hessian. But the sparsity constraints ensure that
the following identities hold for all complex $x$:
\begin{align}
  \Imag(\parD{}{1}{f(x_1,x_2+i\delta)}) -\Imag(\parD{}{1}{f(x_1,x_2)})&=0,\label{eq:I1c}\\
    \Imag(\parD{}{2}{f(x_1+i\delta,x_2)})-\Imag(\parD{}{2}{f(x_1,x_2)})&=0.\label{eq:I2c}
\end{align}
As with the finite difference method, because Equation~\ref{eq:I1c} holds
for $x_1$, it must also hold for $x_1+i\delta$, and because Equation~\ref{eq:I2c} holds
for $x_2$, it must also hold for $x_2+i\delta$. Thus,
\begin{align*}
  \hess{}{f(x)}&\approx
  \begin{pmatrix}
    \Imag(\parD{}{1}{f(x_1+i\delta, x_2+i\delta)})&0\\
        0&\Imag(\parD{}{2}{f(x_1+i\delta,x_2+i\delta)})
    \end{pmatrix}/\delta
\end{align*}
for real $x$.  Only one evaluation of the gradient is required.

@@

Next, define $\Mat{Y}\in\Real{M\times C}$ such that each column is
either a difference in gradients, or the imaginary part of a
complex-valued gradient, depending on the chosen method:
\begin{align}
  \label{eq:Yg}
  \Mat{Y}_c&=
           \begin{cases}
             \nabla f(x+G_c)-\nabla f(x)&\text{finite difference method},\\
             \Imag(\nabla f(x+iG_c))&\text{complex step method}.
             \end{cases}
\end{align}
If $C=M$, then $\Mat{G}$ is a diagonal matrix with $\delta$ in each
diagonal element.  The matrix equation $\hess{}{}\Mat{G}=\Mat{Y}$ represents the linear approximation
$\hess{im}{}\delta\approx y_{im}$, and we can solve for all elements of $\hess{}{}$
just by computing $\Mat{Y}$. But if $C<M$,
there must be at least one $\Mat{G}_c$ with $\delta$ in at least two
rows. The corresponding column $\Mat{Y}_c$ is computed by perturbing multiple variables at
once, so we cannot solve for any $\hess{im}{}$
without further constraints.

These constraints come from the sparsity pattern and symmetry
of the Hessian. Consider an example with the following values and
sparsity pattern:
\begin{align*}
 \hess{}{f(x)}= \begin{pmatrix}
    h_{11}&0&h_{31}&0&0\\
    0&h_{22}&0&h_{42}&0\\
    h_{31}&0&h_{33}&0&h_{53}\\
    0&h_{42}&0&h_{44}&0\\
    0&0&h_{53}&0&h_{55}
  \end{pmatrix}.
\end{align*}
Suppose $C=2$, and define group membership of the five variables through the following $\Mat{G}$ matrix:
\begin{align*}
 \Mat{G}^\top= \begin{pmatrix}
 \delta&\delta&0&0&\delta\\
 0&0&\delta&\delta&0
 \end{pmatrix}.
\end{align*}
Variables 1, 2 and 5 are in group 1, while variables 3 and 4 are in
group 2.
